---
title: "Fetch and Clean 'isd_history.csv' File"
author: "Adam H. Sparks"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The isd_history.csv file details GSOD station metadata.
These data include the start and stop years used by _GSODR_ to pre-check requests before querying the server for download and the country code used by _GSODR_ when sub-setting for requests by country.
The following checks are performed on the raw data file before inclusion in _GSODR_,

  * Check for valid lon and lat values;
    
      * isd_history where latitude or longitude are `NA` or both 0 are removed leaving only properly georeferenced stations,

      * isd_history where latitude is < -90˚ or > 90˚ are removed,

      * isd_history where longitude is < -180˚ or > 180˚ are removed.

  * A new field, STNID, a concatenation of the USAF and WBAN fields, is added.

# Data Processing

## Set up workspace

```{r set_up_workspace, echo=TRUE, message=FALSE, output=FALSE, warning=FALSE}
if (!require("pacman")) {
  install.packages("pacman", repos = "https://cran.rstudio.com/")
}
pacman::p_load("sessioninfo", "skimr", "countrycode", "data.table")
```

## Download and clean data

```{r download_NE_data, echo=TRUE, message=FALSE, warning=FALSE}
# download data
isd_history <- fread("https://www1.ncdc.noaa.gov/pub/data/noaa/isd-history.csv")
```

## Add/drop columns and save to disk

```{r clean_and_reformat, echo=TRUE, cache=FALSE}
# add STNID column
isd_history[, STNID := paste(USAF, WBAN, sep = "-")]
setcolorder(isd_history, "STNID")
setnames(isd_history, "STATION NAME", "NAME")

# drop stations not in GSOD data
isd_history[, STNID_len := nchar(STNID)]
isd_history <- subset(isd_history, STNID_len == 12)

# remove stations where LAT or LON is NA
isd_history <- na.omit(isd_history, cols = c("LAT", "LON"))

# remove extra columns
isd_history[, c("USAF", "WBAN", "ICAO", "ELEV(M)", "STNID_len") := NULL]
```

## Add country names based on FIPS

```{r merge-country, eval=TRUE, message=FALSE}
isd_history <-
  isd_history[setDT(countrycode::codelist), on = c("CTRY" = "fips")]

isd_history <- isd_history[, c(
  "STNID",
  "NAME",
  "LAT",
  "LON",
  "CTRY",
  "STATE",
  "BEGIN",
  "END",
  "country.name.en",
  "iso2c",
  "iso3c"
)]

# clean data
isd_history[isd_history == -999] <- NA
isd_history[isd_history == -999.9] <- NA
isd_history <- isd_history[!is.na(isd_history$LAT) & !is.na(isd_history$LON), ]
isd_history <- isd_history[isd_history$LAT != 0 & isd_history$LON != 0, ]
isd_history <- isd_history[isd_history$LAT > -90 & isd_history$LAT < 90, ]
isd_history <- isd_history[isd_history$LON > -180 & isd_history$LON < 180, ]

# set colnames to upper case
names(isd_history) <- toupper(names(isd_history))
setnames(
  isd_history,
  old = "COUNTRY.NAME.EN",
  new = "COUNTRY_NAME"
)

# set country names to be upper case for easier internal verifications
isd_history[, COUNTRY_NAME := toupper(COUNTRY_NAME)]

# set key for joins when processing CSV files
setkeyv(isd_history, "STNID")
```

## View and save the data

```{r view-and-save}
str(isd_history)

# write rda file to disk for use with GSODR package
save(isd_history,
     file = "../inst/extdata/isd_history.rda",
     compress = "bzip2")
```

# Notes

## NOAA policy

Users of these data should take into account the following (from the
[NCEI website](https://www7.ncdc.noaa.gov/CDO/cdoselect.cmd?datasetabbv=GSOD&countryabbv=&georegionabbv=)): 

> "The following data and products may have conditions placed on their
international commercial use. They can be used within the U.S. or for
non-commercial international activities without restriction. The non-U.S. data
cannot be redistributed for commercial purposes. Re-distribution of these data
by others must provide this same notification."
[WMO Resolution 40. NOAA Policy](http://www.wmo.int/pages/about/Resolution40.html)

## R System Information

```{r system_information, echo=FALSE}
session_info()
```
