---
title: "Fetch, clean and correct altitude in GSOD isd_history.csv data"
author: "Adam H. Sparks"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This document details how the GSOD station history data file,
["isd-history.csv"](ftp://ftp.ncdc.noaa.gov/pub/data/noaa/isd-history.csv),
is fetched from the NCEI ftp server, error checked and new elevation values
generated. The new elevation values are then saved for inclusion in package as
`/extdata/isd_history.rda`. The resulting values are merged with the most
recent station history data file from the NCEI when the user runs the
`get_GSOD()` function. The resulting data frame of station information, based on
the merging of the `corrected_elev` data frame with the most recently
available "isd-history.csv" file, will result in the following changes to the
data:

  * Stations where latitude or longitude are NA or both 0 are removed

  * Stations where latitude is < -90˚ or > 90˚ are removed

  * Stations where longitude is < -180˚ or > 180˚ are removed

  * A new field, STNID, a concatenation of the USAF and WBAN fields, is added

  * Stations are checked against Natural Earth 1:10 ADM0 cultural data, stations
not mapping in the isd-history reported country are dropped

  * 90m hole-filled SRTM digital elevation (Jarvis *et al.* 2008) is used to
identify and correct/remove elevation errors in data for station locations
between -60˚ and 60˚ latitude. This applies to cases here where elevation was
missing in the reported values as well. In case the station reported an
elevation and the DEM does not, the station reported value is taken. For
stations beyond -60˚ and 60˚ latitude, the values are station reported values in
every instance for the 90m column.

# Data Processing

## Set up workspace

```{r set_up_workspace, echo=TRUE, messages=FALSE, output=FALSE, warning=FALSE}
if (!require("dplyr")) {
  install.packages("dplyr", repos = "https://cran.rstudio.com/")
}

if (!require("doParallel")) {
  install.packages("doParallel", repos = "https://cran.rstudio.com/")
}

if (!require("foreach")) {
  install.packages("foreach", repos = "https://cran.rstudio.com/")
}

if (!require("ggplot2")) {
  install.packages("ggplot2", repos = "https://cran.rstudio.com/")
}

if (!require("parallel")) {
  install.packages("parallel", repos = "https://cran.rstudio.com/")
}

if (!require("raster")) {
  install.packages("raster", repos = "https://cran.rstudio.com/")
}

if (!require("readr")) {
  install.packages("readr", repos = "https://cran.rstudio.com/")
}

if (!require("rnaturalearth")) {
  install.packages("rnaturalearth", repos = "https://cran.rstudio.com/")
}

if (!require("hrbrthemes")) {
  install.packages("hrbrthemes", repos = "https://cran.rstudio.com/")
}

library(magrittr) # comes with dplyr above

dem_tiles <- list.files(path.expand("~/Data/CGIAR-CSI SRTM"), 
                        pattern = glob2rx("*.tif"), full.names = TRUE)
crs <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"
cor_stations <- list()
tf <- tempfile()
```

## Download data from Natural Earth and NCEI

```{r download_NE_data, echo=TRUE, message=FALSE, warning=FALSE}
# import Natural Earth cultural 1:10m data
NE <- rnaturalearth::ne_countries(scale = 10)

# download data
stations <- readr::read_csv(
  "ftp://ftp.ncdc.noaa.gov/pub/data/noaa/isd-history.csv",
  col_types = "ccccccddddd",
  col_names = c("USAF", "WBAN", "STN_NAME", "CTRY", "STATE", "CALL",
                "LAT", "LON", "ELEV_M", "BEGIN", "END"), skip = 1)

stations[stations == -999] <- NA
stations[stations == -999.9] <- NA
```

## Reformat and clean station data file from NCEI

```{r clean_and_reformat, echo=TRUE, cache=FALSE}
# clean data
stations <- stations[!is.na(stations$LAT) & !is.na(stations$LON), ]
stations <- stations[stations$LAT != 0 & stations$LON != 0, ]
stations <- stations[stations$LAT > -90 & stations$LAT < 90, ]
stations <- stations[stations$LON > -180 & stations$LON < 180, ]
stations$STNID <- as.character(paste(stations$USAF, stations$WBAN, sep = "-"))
```

## Check data for inconsistencies

### Check for country of station location

GSOD data have some inconsistencies in them, some of this has been removed
above with filtering. Further filtering is used remove stations reporting
locations in countries that do not match the physical coordinates reported.
Using
[Natural Earth Data 1:10 Cultural Data](http://www.naturalearthdata.com/downloads/10m-cultural-vectors/),
the stations reported countries are checked against the country in which the
coordinates map.

Also, reported elevation may differ from actual. Hijmans *et al.* (2005) created
their own digital elevation model using Jarvis *et al.* (2004) and 
[GTOPO30 data](https://lta.cr.usgs.gov/GTOPO30)
for areas where there was no SRTM data available (>+/-60˚ latitude). Here only
the hole-filled SRTM data, V4 (Jarvis *et al.* 2008) was used for correction of
agroclimatology data (-60˚ to 60˚ latitude). Any incorrect station elevations
beyond these values were ignored in this data set. Stations with incorrect
elevation were identified using
`raster::extract(x, y, buffer = 200, fun = mean)` so that surrounding cells are
also used to determine the elevation at that point, reducing the chances of over
or underestimating in mountainous areas. See Hijmans *et al.* (2005) for more 
detailed information on this methodology.

The hole-filled SRTM data is large enough that it won't all fit in-memory on
most desktop computers. Using tiles allows this process to run on a modest
machine with minimal effort but does take some time to loop through all of the
tiles.

Data can be downloaded from the
[CGIAR-CSI's](http://csi.cgiar.org/WhtIsCGIAR_CSI.asp) ftp server, [srtm.csi.cgiar.org](ftp://srtm.csi.cgiar.org), using an FTP client to
facilitate this next step.

```{r check_station_location, echo=TRUE, cache=FALSE}
# quality check station locations for reported country and lat/lon position
# agreement

# create spatial object to check for location
xy <- as.data.frame(stations)
sp::coordinates(xy) <- ~ LON + LAT
sp::proj4string(xy) <- sp::CRS(crs)

# check for location in country
point_check <- sp::over(xy, NE)
point_check <- as.data.frame(point_check)
stations_discard <- point_check[point_check$FIPS %in% point_check$FIPS_10_ == FALSE, ]
nrow(stations_discard)
```
Zero observations (rows) in `stations_discard`, the data look good, no need to
remove any

### Elevation data supplement

Next use the `raster::extract()` function to get the mean elevation data from
the 90m elevation data and supplement the elevation data from the NCEI.

```{r check_and_supplement_elevation, echo=TRUE, cache=FALSE}
# create a spatial object for extracting elevation values using spatial points
stations <- as.data.frame(stations)
sp::coordinates(stations) <- ~ LON + LAT
sp::proj4string(stations) <- sp::CRS(crs)

# set up cluster for parallel processing
library(foreach)
cl <- parallel::makeCluster(parallel::detectCores())
doParallel::registerDoParallel(cl)

corrected_elev <- dplyr::bind_rows(
  foreach(i = dem_tiles, .packages = "magrittr") %dopar% {
    # Load the DEM tile
    dem <- raster::raster(i)
    sub_stations <- raster::crop(stations, dem)
    
    # in some cases the DEM represents areas where there is no station
    # check for that here and if no stations, go on to next iteration
    if (!is.null(sub_stations)) {
      # use a 200m buffer to extract elevation from the DEM
      
      sub_stations$ELEV_M_SRTM_90m <- 
        raster::extract(dem, sub_stations,
                        buffer = 200,
                        fun = mean)
      
      # convert spatial object back to normal data frame and add new fields
      sub_stations <- as.data.frame(sub_stations)
      
      # set any factors back to character
      sub_stations <- sub_stations %>%
        dplyr::mutate_if(is.factor, as.character)
      
      return(sub_stations)
    }
  }
)

# stop cluster
parallel::stopCluster(cl)
```
Some stations occur in areas where DEM has no data, in this case, use original
station elevation for these stations.

```{r create corrected_elev}
corrected_elev <- dplyr::mutate(corrected_elev,
                                ELEV_M_SRTM_90m = ifelse(
                                  is.na(ELEV_M_SRTM_90m),
                                  ELEV_M, ELEV_M_SRTM_90m))
```

In some cases duplicate stations occur, use the mean value of duplicate rows for
corrected elevation and create a data frame with only `STNID` and the new
elevation values. `STNID` is used for a left-join with the `stations` object.

```{r summarise_corrected}
corrected_elev <- corrected_elev %>%
  dplyr::group_by(STNID) %>%
  dplyr::summarise(ELEV_M_SRTM_90m = mean(ELEV_M_SRTM_90m))
```

Round `ELEV_M_SRTM_90m` field to whole number in cases where station reported
data was used.

```{r round_corrected}
corrected_elev[, 2] <- round(corrected_elev[, 2], 0)
```

Left-join the new station elevation data with the `stations` object. For
stations above/below 60/-60 latitude or buoys, `ELEV_M_SRTM_90m` will be `NA` as
there is no SRTM data for these locations.

```{r join_corrected_elev}
# convert any factors in stations object to character for left_join
stations <- dplyr::mutate_if(
  as.data.frame(stations),
  is.factor,
  as.character)

# Perform left join to join corrected elevation with original station data,
# this will include stations below/above -60/60
isd_history <- 
  dplyr::left_join(stations, corrected_elev,
                   by = "STNID") %>% 
  tibble::as_tibble() %>% 
  dplyr::mutate(BEGIN = as.integer(BEGIN),
                END = as.integer(END))

str(isd_history)

isd_history
```

# Figures

Visualise the corrected elevation values against the original elevation values.

``````{r Buffered SRTM 90m vs Reported Elevation, echo=TRUE, fig.cap="GSOD Reported Elevation versus CGIAR-CSI SRTM Buffered Elevation", warning=FALSE}

ggplot2::ggplot(data = isd_history,
                ggplot2::aes(x = ELEV_M,
                             y = ELEV_M_SRTM_90m)) +
  ggplot2::geom_point(alpha = 0.4,
                      size = 0.5) +
  ggplot2::geom_abline(slope = 1,
                       colour = "white",
                       width = 2.5) +
  ggplot2::ggtitle("Corrected elevation versus original elevation values.") +
  hrbrthemes::theme_ipsum()
  
```

Buffered versus non-buffered elevation values were previously checked and found
not to be different while also not showing any discernible geographic patterns.
However, the buffered elevation data are higher than the non-buffered data. To
help avoid within cell and between cell variation the buffered values are the
values that are included in the final data for distribution with the GSODR
package following the approach of Hijmans *et al.* (2005).

The final dataframe for distribution with _GSODR_ includes the new elevation 
values in the `ELEV\_M\_SRTM\_90m` field along with the cleaned
"isd-history.csv" data that removes stations that do not have valid x, y
locations.

```{r cleanup_and_.rda_file_creation, message=FALSE, echo=TRUE, warning=FALSE, error=FALSE}

# write rda file to disk for use with GSODR package
save(isd_history, file = "../inst/extdata/isd_history.rda",
     compress = "bzip2")
```

# Notes

## NOAA Policy

Users of these data should take into account the following (from the
[NCEI website](http://www7.ncdc.noaa.gov/CDO/cdoselect.cmd?datasetabbv=GSOD&countryabbv=&georegionabbv=)): 

> "The following data and products may have conditions placed on their
international commercial use. They can be used within the U.S. or for
non-commercial international activities without restriction. The non-U.S. data
cannot be redistributed for commercial purposes. Re-distribution of these data
by others must provide this same notification."
[WMO Resolution 40. NOAA Policy](http://www.wmo.int/pages/about/Resolution40.html)

## R System Information

```{r system_information, echo=FALSE}
sessioninfo::session_info()
```

# References

Hijmans, RJ, SJ Cameron, JL Parra, PG Jones, A Jarvis, 2005, Very High
Resolution Interpolated Climate Surfaces for Global Land Areas. *International
Journal of Climatology*. 25: 1965-1978.
[DOI:10.1002/joc.1276](http://dx.doi.org/10.1002/joc.1276)

Jarvis, A, HI Reuter, A Nelson, E Guevara, 2008, Hole-filled SRTM for the
globe Version 4, available from the CGIAR-CSI SRTM 90m Database
([http://srtm.csi.cgiar.org](http://srtm.csi.cgiar.org))

Jarvis, A, J Rubiano, A Nelson, A Farrow and M Mulligan, 2004, Practical use of
SRTM Data in the Tropics: Comparisons with Digital Elevation Models Generated
From Cartographic Data. Working Document no. 198. Cali, CO.
International Centre for Tropical Agriculture (CIAT): 32.
[URL](http://srtm.csi.cgiar.org/PDF/Jarvis4.pdf)
