---
title: "Fetch GSOD Country List and Merge with ISO Country Codes"
author: "Adam H. Sparks - Center for Crop Health, University of Southern Queensland"
date: "07-011-2016"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This script will fetch the country list provided by the NCDC for the GSOD
stations from the ftp server and merge it with ISO codes from the
[`countrycode`](https://github.com/vincentarelbundock/countrycode)
package for inclusion in the GSODR package in /data/country-list.rda. These
codes are used when a user selects a single country for a data query.

This inclusion decreases the time necessary to query the server when specifying
a country for weather data downloading.

# R Data Processing

Read "country-list.txt"" file from NCDC FTP server and merge with`countrycode`
data.

```{r download_merge_inspect, echo=TRUE, messages=FALSE}
countries <- readr::read_table(
  "ftp://ftp.ncdc.noaa.gov/pub/data/noaa/country-list.txt")[-1, c(1, 3)]

country_list <- dplyr::left_join(countries, countrycode::countrycode_data,
                              by = c(FIPS = "fips104"))

print(country_list)
```

There are unecessary data in several columns. `GSODR` only requires FIPS, name,
and ISO codes to function.

```{r remove_uncessesary, echo=TRUE, messages=FALSE}
country_list <- country_list[, -c(3, 4:8, 11:16)]

print(country_list)
```

Write .rda file to disk for inclusion in `GSODR` package.

```{r write, echo=TRUE, messages=FALSE}
devtools::use_data(country_list, overwrite = TRUE, compress = "bzip2")
```
