---
title: "Fetch, clean and correct altitude in GSOD isd_history.csv Data"
author: "Adam H. Sparks"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This script will fetch station data from the ftp server and clean up for
inclusion in package in /data/stations.rda for the GSODR package.

The following changes are made:

* Stations where latitude or longitude are NA or both 0 were removed

* Stations where latitude is < -90˚ or > 90˚ were removed

* Stations where longitude is < -180˚ or > 180˚ were removed

* A new field, STNID, a concatenation of the USAF and WBAN fields, was added

* Stations were checked against Natural Earth 1:10 ADM0 Cultural data, stations
not mapping in the isd-history reported country were dropped

* 90m hole-filled SRTM digital elevation (Jarvis *et al.* 2008) was used to
identify and correct/remove elevation errors in data for station locations
between -60˚ and 60˚ latitude. This applies to cases here where elevation was
missing in the reported values as well. In case the station reported an
elevation and the DEM does not, the station reported value is taken. For
stations beyond -60˚ and 60˚ latitude, the values are station reported values in
every instance for the 90m column.

# Data Processing

## Set up workspace

```{r set up workspace, echo=TRUE, messages=FALSE}
dem_tiles <- list.files(path.expand("~/Data/CGIAR-CSI SRTM"), 
                        pattern = glob2rx("*.tif"), full.names = TRUE)
crs <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"
cor_stations <- list()
tf <- tempfile()
```

## Download from Natural Earth and NCDC

```{r download NE data, echo=TRUE}
# import Natural Earth cultural 1:10m data
curl::curl_download("http://www.naturalearthdata.com/http//www.naturalearthdata.com/download/10m/cultural/ne_10m_admin_0_countries.zip",
                    destfile = tf)
NE <- unzip(tf)
NE <- raster::shapefile("./ne_10m_admin_0_countries.shp")
unlink(tf)

# download data
stations <- readr::read_csv(
  "ftp://ftp.ncdc.noaa.gov/pub/data/noaa/isd-history.csv",
  col_types = "ccccccddddd",
  col_names = c("USAF", "WBAN", "STN_NAME", "CTRY", "STATE", "CALL",
                "LAT", "LON", "ELEV_M", "BEGIN", "END"), skip = 1)

stations[stations == -999.9] <- NA
stations[stations == -999] <- NA

countries <- readr::read_table(
  "ftp://ftp.ncdc.noaa.gov/pub/data/noaa/country-list.txt")[-1, c(1, 3)]

```

## Reformat and clean station data file from NCDC

```{r clean and reformat station data, echo=TRUE, cache=FALSE}
# clean data
stations <- stations[!is.na(stations$LAT) & !is.na(stations$LON), ]
stations <- stations[stations$LAT != 0 & stations$LON != 0, ]
stations <- stations[stations$LAT > -90 & stations$LAT < 90, ]
stations <- stations[stations$LON > -180 & stations$LON < 180, ]
stations$STNID <- as.character(paste(stations$USAF, stations$WBAN, sep = "-"))

xy <- dplyr::left_join(stations, countries, by = c("CTRY" = "FIPS"))

# join with countrycode data
countries <- dplyr::left_join(countries, countrycode::countrycode_data,
                              by = c(FIPS = "fips104"))
```

## Check data for inconsistencies

GSOD data have some inconsistencies in them, some of this has been removed
above with filtering. Further filtering is used remove stations reporting
locations in countries that do not match the physical coordinates reported.
Using [Natural Earth Data 1:10 Cultural Data](http://www.naturalearthdata.com/downloads/10m-cultural-vectors/),
the stations reported countries are checked against the country in which the
coordinates map.

Also, reported elevation may differ from actual. Hijmans *et al.* (2005) created
their own digital elevation model using Jarvis *et al.* (2004) and [GTOPO30 data](https://lta.cr.usgs.gov/GTOPO30)
for areas where there was no SRTM data available (>+/-60˚ latitude). Here only
the hole-filled SRTM data, V4 (Jarvis *et al.* 2008) was used for correction of
agroclimatology data (-60˚ to 60˚ latitude). Any incorrect station elevations
beyond these values were ignored in this data set. Stations with incorrect
elevation were identified using
`raster::extract(x, y, buffer = 200, fun = mean)` so that surrounding cells are
also used to determine the elevation at that point, reducing the chances of over
or underestimating in mountainous areas. See Hijmans *et al.* (2005) for more 
detailed information on this methodology.

The hole-filled SRTM data is large enough that it won't all fit in-memory on
most desktop computers. Using tiles allows this process to run on a modest
machine with minimal effort but does take some time to loop through all of the
tiles.

Data can be downloaded from the
[CGIAR-CSI's](http://csi.cgiar.org/WhtIsCGIAR_CSI.asp)
ftp server, [srtm.csi.cgiar.org](ftp://srtm.csi.cgiar.org),
using an FTP client to facilitate this next step.

```{r check station data for errors, echo=TRUE, cache=FALSE}
# quality check station locations for reported country and lat/lon position
# agreement

# create spatial object to check for location
xy <- as.data.frame(xy)
sp::coordinates(xy) <- ~ LON + LAT
sp::proj4string(xy) <- sp::CRS(crs)

# check for location in country
point_check <- sp::over(xy, NE)
point_check <- as.data.frame(point_check)
stn_discard <- point_check[point_check$FIPS %in% point_check$FIPS_10_ == FALSE, ]
nrow(stn_discard)
```
Zero observations (rows) in stations_discard, the data look good, no need to
remove any


```{r check elevation errors, echo=TRUE, cache=FALSE}
# create a spatial object for extracting elevation values using spatial points
stations <- as.data.frame(stations)
sp::coordinates(stations) <- ~ LON + LAT
sp::proj4string(stations) <- sp::CRS(crs)

# set up cluster for parallel processing
library(foreach)
library(parallel)
library(doParallel)
cl <- makeCluster(detectCores() - 2)
registerDoParallel(cl)

corrected_elev <- tibble::as_tibble(data.table::rbindlist(foreach(i = dem_tiles) %dopar% {
  
  # Load the DEM tile
  dem <- raster::raster(i)
  cropped_stations <- raster::crop(stations, dem)
  
  # in some cases the DEM represents areas where there is no station
  # check for that here and if no stations, go on to next iteration
  if (!is.null(cropped_stations)) {
    
    # use a 200m buffer to extract elevation from the DEM
    SRTM_buffered <- raster::extract(dem, cropped_stations, buffer = 200, fun = mean)

    # convert spatial object back to normal data frame and add new fields
    corrected_elev <- tibble::as_tibble(cropped_stations)
    corrected_elev$ELEV_M_SRTM_90m_BUFFER <- SRTM_buffered
    return(corrected_elev)
  }
}
)
)

# stop cluster
stopCluster(cl)
```
Some stations occur in areas where DEM has no data, in this case, use original
station elevation for these stations

```{r}
# convert stations from a spatial object to a tible for joining

stations <- tibble::as_tibble(stations)
corrected_elev[, 13] <- ifelse(is.na(corrected_elev[, 13]),
                               corrected_elev[, 9], corrected_elev[, 13])

# Perform left join to join corrected elevation with original station data,
# this will include stations below/above -60/60
test <- dplyr::left_join(stations, corrected_elev)

summary(stations)
```

# Figures

```{r ggplot, echo=FALSE, message=FALSE}
library(ggplot2)
library(ggalt)
```

```{r Buffered SRTM 90m vs Reported Elevation, echo=TRUE, fig.cap="GSOD Reported Elevation versus CGIAR-CSI SRTM Buffered Elevation", warning=FALSE}
ggplot(data = stations, aes(x = ELEV_M, y = ELEV_M_SRTM_90m_BUFFER)) +
  geom_point(alpha = 0.4, size = 0.5)
```

Buffered versus unbuffered elevation values were previously checked and found
not to be different while also not showing any discernable geographic patterns.
However, The buffered elevation data are higher than the unbuffered data. To
help avoid within cell and between cell variation the buffered values are the
values that are included in the final data for distribution with the GSODR
package following the approach of Hijmans *et al.* (2005).

```{r cleanup and .rda file creation, message=FALSE, echo=TRUE, warning=FALSE, error=FALSE}

# round SRTM_90m_Buffer field to whole number in cases where station reported
# data was used and rename column
stations[, 13] <- round(stations[, 13], 0)
names(stations)[13] <- "ELEV_M_SRTM_90m"

# Set STNID column to character
stations[, 12] <- as.character(stations[, 12])

# convert to regular data.frame object and write rda file to disk
stations <- data.frame(stations)
devtools::use_data(stations, overwrite = TRUE, compress = "bzip2")

# clean up Natural Earth data files before we leave
file.remove(list.files(pattern = glob2rx("ne_10m_admin_0_countries*")))

```

The stations.rda file included in the GSODR package includes
the new elevation data as the field; ELEV\_M\_SRTM\_90m.

# Notes

## NOAA Policy

Users of these data should take into account the following (from the [NCDC website](http://www7.ncdc.noaa.gov/CDO/cdoselect.cmd?datasetabbv=GSOD&countryabbv=&georegionabbv=)): 

> "The following data and products may have conditions placed on their international commercial use. They can be used within the U.S. or for non-commercial international activities without restriction. The non-U.S. data cannot be redistributed for commercial purposes. Re-distribution of these data by others must provide this same notification." [WMO Resolution 40. NOAA Policy](http://www.wmo.int/pages/about/Resolution40.html)

## R System Information

```{r system information, echo=FALSE}
sessionInfo()
```

# References

Hijmans, RJ, SJ Cameron, JL Parra, PG Jones, A Jarvis, 2005, Very High
Resolution Interpolated Climate Surfaces for Global Land Areas. *International
Journal of Climatology*. 25: 1965-1978.
[DOI:10.1002/joc.1276](http://dx.doi.org/10.1002/joc.1276)

Jarvis, A, HI Reuter, A Nelson, E Guevara, 2008, Hole-filled SRTM for the
globe Version 4, available from the CGIAR-CSI SRTM 90m Database
([http://srtm.csi.cgiar.org](http://srtm.csi.cgiar.org))

Jarvis, A, J Rubiano, A Nelson, A Farrow and M Mulligan, 2004, Practical use of
SRTM Data in the Tropics: Comparisons with Digital Elevation Models Generated
From Cartographic Data. Working Document no. 198. Cali, CO.
International Centre for Tropical Agriculture (CIAT): 32. [URL](http://srtm.csi.cgiar.org/PDF/Jarvis4.pdf)

