---
title: "Fetch GSOD Country List and Merge with ISO Country Codes"
author: "Adam H. Sparks"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Introduction

This document details how to fetch the country list provided by the NCEI for
the GSOD stations from the FTP server and merge it with ISO codes from the
[_countrycode_](https://cran.r-project.org/package=countrycode)
package for inclusion in the _GSODR_ package in `/data/country-list.rda`. These
codes are used when a user selects a single country for a data query.

# R Data Processing

Read "country-list.txt" file from NCEI FTP server and merge with _countrycode_
data.

```{r download-merge-inspect, echo=TRUE, messages=FALSE}
if (!require("countrycode"))
{
  install.packages("countrycode",
                   repos = c(CRAN = "https://cran.rstudio.com"))
}
if (!require("dplyr"))
{
  install.packages("dplyr",
                   repos = c(CRAN = "https://cran.rstudio.com"))
}
```

```{r read-table, echo=TRUE, messages=FALSE}

```{r download-and-merge, echo=TRUE, message=FALSE}
countries <- readr::read_table(
  "ftp://ftp.ncdc.noaa.gov/pub/data/noaa/country-list.txt")[-1, c(1, 3)]
names(countries)[2] <- "COUNTRY_NAME"

country_list <- dplyr::left_join(countries, countrycode::codelist,
                   by = c("FIPS" = "fips"))

print(country_list)
```

There are unnecessary data in several columns. _GSODR_ only requires FIPS, name,
and ISO codes to function.


```{r remove-uncessesary, echo=TRUE, messages=FALSE}
country_list <- dplyr::select(country_list, c("FIPS",
                                              "COUNTRY_NAME",
                                              "iso2c",
                                              "iso3c"))
country_list
```

Write .rda file to disk.

```{r write, echo=TRUE, messages=FALSE}
save(country_list, file = "../inst/extdata/country_list.rda",
     compress = "bzip2")
```
# Notes

## NOAA Policy

Users of these data should take into account the following (from the 
[NCEI website](http://www7.ncdc.noaa.gov/CDO/cdoselect.cmd?datasetabbv=GSOD&countryabbv=&georegionabbv=)): 
> "The following data and products may have conditions placed on their international commercial use. They can be used within the U.S. or for
> non-commercial international activities without restriction. The non-U.S.
> data cannot be redistributed for commercial purposes. Re-distribution of these
> data by others must provide this same notification."
[WMO Resolution 40. NOAA Policy](http://www.wmo.int/pages/about/Resolution40.html)

## R System Information

```{r system information, echo=FALSE}
sessioninfo::session_info()
```
