---
title: "CRU CL2.0 Data"
output: github_document
---

```{r knitr_setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

CRU CL2.0 data are a gridded climatology of 1961-1990 monthly means released in 2002 and cover all land areas (excluding Antarctica) at 10-minute resolution. For more information see the description of the data provided by the University of East Anglia Climate Research Unit (CRU), http://www.cru.uea.ac.uk/cru/data/hrg/tmc/readme.txt.

# Download, extract and merge CRU data with provided GSOD climate data

## Setup the R session

```{r session_setup}
library(getCRUCLdata)

# load existing GSOD_clim data from package
GSOD_clim <- GSODR::GSOD_clim
```
## Get CRU CL2.0 data
```{r get_data}
CRU_stack <- create_CRU_stack(pre = TRUE,
                              rd0 = TRUE,
                              tmp = TRUE,
                              dtr = TRUE,
                              reh = TRUE,
                              sunp = TRUE,
                              frs = TRUE,
                              wnd = TRUE)
```

## Extract data for station locations
```{r extract_data}
stations <- readr::read_csv(
  "ftp://ftp.ncdc.noaa.gov/pub/data/noaa/isd-history.csv",
  col_types = "ccccccddddd",
  col_names = c("USAF", "WBAN", "STN_NAME", "CTRY", "STATE", "CALL",
                "LAT", "LON", "ELEV_M", "BEGIN", "END"), skip = 1)

stations[stations == -999.9] <- NA
stations[stations == -999] <- NA
stations <- stations[!is.na(stations$LAT) & !is.na(stations$LON), ]
stations <- stations[stations$LAT != 0 & stations$LON != 0, ]
stations <- stations[stations$LAT > -90 & stations$LAT < 90, ]
stations <- stations[stations$LON > -180 & stations$LON < 180, ]
stations <- stations[!is.na(stations$STN_NAME), ]
stations$STNID <- as.character(paste(stations$USAF, stations$WBAN, sep = "-"))

stations <- as.data.frame(stations)
sp::coordinates(stations) <- ~ LON + LAT
crs <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"
sp::proj4string(stations) <- sp::CRS(crs)

# create a vector of names for the raster layers in new stack
CRU_stack_names <- c(
  paste0(
    "CRU_CL2_0_",
    names(CRU_stack[1]),
    "_",
    c(
      "01",
      "02",
      "03",
      "04",
      "05",
      "06",
      "07",
      "08",
      "09",
      "10",
      "11",
      "12"
    )
  ),
  paste0(
    "CRU_CL2_0_",
    names(CRU_stack[2]),
    "_",
    c(
      "01",
      "02",
      "03",
      "04",
      "05",
      "06",
      "07",
      "08",
      "09",
      "10",
      "11",
      "12"
    )
  ),
  paste0(
    "CRU_CL2_0_",
    names(CRU_stack[3]),
    "_",
    c(
      "01",
      "02",
      "03",
      "04",
      "05",
      "06",
      "07",
      "08",
      "09",
      "10",
      "11",
      "12"
    )
  ),
  paste0(
    "CRU_CL2_0_",
    names(CRU_stack[4]),
    "_",
    c(
      "01",
      "02",
      "03",
      "04",
      "05",
      "06",
      "07",
      "08",
      "09",
      "10",
      "11",
      "12"
    )
  ),
  paste0(
    "CRU_CL2_0_",
    names(CRU_stack[5]),
    "_",
    c(
      "01",
      "02",
      "03",
      "04",
      "05",
      "06",
      "07",
      "08",
      "09",
      "10",
      "11",
      "12"
    )
  ),
  paste0(
    "CRU_CL2_0_",
    names(CRU_stack[6]),
    "_",
    c(
      "01",
      "02",
      "03",
      "04",
      "05",
      "06",
      "07",
      "08",
      "09",
      "10",
      "11",
      "12"
    )
  ),
  paste0(
    "CRU_CL2_0_",
    names(CRU_stack[7]),
    "_",
    c(
      "01",
      "02",
      "03",
      "04",
      "05",
      "06",
      "07",
      "08",
      "09",
      "10",
      "11",
      "12"
    )
  ),
  paste0(
    "CRU_CL2_0_",
    names(CRU_stack[8]),
    "_",
    c(
      "01",
      "02",
      "03",
      "04",
      "05",
      "06",
      "07",
      "08",
      "09",
      "10",
      "11",
      "12"
    )
  )
)

# Create one stack object from list of stacks
CRU_stack <-  raster::stack(unlist(CRU_stack))
names(CRU_stack) <- CRU_stack_names

# Extract CRU data at GSOD station locations
CRU_GSOD <- raster::extract(CRU_stack, stations)
CRU_GSOD <- data.frame(as.data.frame(stations$STNID), CRU_GSOD)
```

## Merge data sets
```{r merge}
# Convert STNID to character
CRU_GSOD$stations.STNID <- as.character(CRU_GSOD$stations.STNID)
GSOD_clim$STNID <- as.character(GSOD_clim$STNID)

# Left join GSOD_clim data with the new CRU data
GSOD_clim <- dplyr::left_join(GSOD_clim, CRU_GSOD, by = c("STNID" = "stations.STNID"))
```

## Save new data to disk for distribution with R package
```{r save}
devtools::use_data(GSOD_clim, overwrite = TRUE, compress = "bzip2")
```

# Data reference and abstract

> Mark New (1,*), David Lister (2), Mike Hulme (3), Ian Makin (4)  
> A high-resolution data set of surface climate over global land areas 
Climate Research, 2000, Vol 21, pg 1-25  
> (1) School of Geography and the Environment, University of Oxford, 
Mansfield Road, Oxford OX1 3TB, United Kingdom  
> (2) Climatic Research Unit, and (3) Tyndall Centre for Climate Change Research,
both at School of Environmental Sciences, University of East Anglia,
Norwich NR4 7TJ, United Kingdom  
> (4) International Water Management Institute, PO Box 2"07"5, Colombo, Sri Lanka

> **ABSTRACT:** We describe the construction of a 10-minute latitude/longitude data
set of mean monthly surface climate over global land areas, excluding Antarctica.
The climatology includes 8 climate elements - precipitation, wet-day frequency,
temperature, diurnal temperature range, relative humidity,sunshine duration,
ground frost frequency and windspeed - and was interpolated from a data set
of station means for the period centred on 1961 to 1990. Precipitation was first
defined in terms of the parameters of the Gamma distribution, enabling the 
calculation of monthly precipitation at any given return period. The data are
compared to an earlier data set at 0.5 degrees latitude/longitude resolution and
show added value over most regions. The data will have many applications in
applied climatology, biogeochemical modelling, hydrology and agricultural
meteorology and are available through the School of Geography Oxford 
(http://www.geog.ox.ac.uk), the International Water Management Institute
"World Water and Climate Atlas" (http://www.iwmi.org) and the Climatic
Research Unit (http://www.cru.uea.ac.uk).


## R System Information

```{r system information, echo=FALSE}
sessionInfo()
```
