---
title: "Fetch, clean and correct altitude in GSOD isd_history.csv Data"
author: "Adam H. Sparks - Center for Crop Health, University of Southern Queensland"
date: "05-30-2016"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This script will fetch station data from the ftp server and clean up for
inclusion in package in /data/stations.rda for the GSODR package.

The following changes are made:

* Stations where latitude or longitude are NA or both 0 were removed

* Stations where latitude is < -90˚ or > 90˚ were removed

* Stations where longitude is < -180˚ or > 180˚ were removed

* A new field, STNID, a concatenation of the USAF and WBAN fields, was added

* Stations were checked against FAO-GAUL 2015 data, countries not mapping
in the isd-history reported country were dropped

* 90m hole-filled SRTM digital elevation (Jarvis *et al.* 2008) was used to
identify and correct/remove elevation errors in dat for station locations
between -60˚ and 60˚. *Only for agroclimatology option data*


# R Data Processing

## Load libraries and set up workspace

```{r, echo=TRUE, messages=FALSE}
dem_tiles <- list.files(path.expand("~/tmp/srtm"), full.names = TRUE)
crs <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"
cor_stations <- list()

```


## Download, reformat and clean station data file from NCDC

```{r, echo=TRUE}

# Load GAUL
GAUL <- raster::shapefile("~/Google Drive/Data/GAUL/g2015_2014_0/g2015_2014_0.shp")

# Download data
stations <- readr::read_csv(
  "ftp://ftp.ncdc.noaa.gov/pub/data/noaa/isd-history.csv",
  col_types = "ccccccddddd",
  col_names = c("USAF", "WBAN", "STN.NAME", "CTRY", "STATE", "CALL",
                "LAT", "LON", "ELEV.M", "BEGIN", "END"), skip = 1,
  na = c("-999.9", "-999.0"))

countries <- readr::read_table(
  "ftp://ftp.ncdc.noaa.gov/pub/data/noaa/country-list.txt")[-1, c(1, 3)]


# Clean data
stations <- stations[!is.na(stations$LAT) & !is.na(stations$LON), ]
stations <- stations[stations$LAT != 0 & stations$LON != 0, ]
stations <- stations[stations$LAT > -90 & stations$LAT < 90, ]
stations <- stations[stations$LON > -180 & stations$LON < 180, ]
stations$STNID <- paste(stations$USAF, stations$WBAN, sep = "-")


countries <- dplyr::left_join(countries, countrycode::countrycode_data,
                              by = c(FIPS = "fips104"))
stations <- dplyr::left_join(stations, countries, by = c("CTRY" = "FIPS"))
stations <- as.data.frame(stations) # convert to dataframe object

# Create spatial object extracting elevation values using spatial points
sp::coordinates(stations) <- ~LON + LAT
sp::proj4string(stations) <- sp::CRS(crs)

```

## Loop to check elevation values for inconsistencies

GSOD data have some inconsistencies in them, some of this has been removed
above with filtering. Further filtering is used remove stations reporting
locations in countries that do not match the physical coordinates reported.
Using [FAO-GAUL (Global Administrative Unit Layers) 2015](http://www.fao.org/geonetwork/srv/en/metadata.show?id=12691),
the stations reported countries are checked against the country in which the
coordinates map.

Also, reported elevation may differ from actual. Hijmans *et al.* (2005) created
their own digital elevation model using Jarvis *et al.* (2004) and [GTOPO30 data](https://lta.cr.usgs.gov/GTOPO30)
for areas where there was no SRTM data available (>60˚). Here only the
hole-filled SRTM data, V4 (Jarvis *et al.* 2008) was used for correction of
agroclimatology data (-60˚ to 60˚). Any incorrect station elevations beyond
these values were ignored in this data set. Stations with incorrect elevation
were identified using `raster::extract(x, y, method = "bilinear")` in order that
values of the four nearest cells are also used to determine elevation errors.
See Hijmans *et al.* (2005) for more detailed information on this.

The hole-filled SRTM data is large enough that it won't all fit in-memory on
most desktop computers. Using tiles allows this process to run on a modest
machine with minimal effort but does take some time to loop through all of the
tiles.

Data can be downloaded from the [CGIAR-CSI's](http://csi.cgiar.org/WhtIsCGIAR_CSI.asp)
ftp server, [srtm.csi.cgiar.org](ftp://srtm.csi.cgiar.org),
using an FTP client to facilitate this next step.

```{r, echo=TRUE}

# Quality check station locations for reported country and lat/lon position
# agreement
point_check <- over(stations, GAUL)
stations <- as.data.frame(stations)

stations_keep <- stations[stations$country.name %in% point_check$ADM0_NAME == TRUE, ]
stations_discard <- stations[stations$country.name %in% point_check$ADM0_NAME == FALSE, ]

for (i in dem_tiles) {
  
  # Load the DEM tile
  dem <- raster::raster(list.files(paste0(i, "/"), pattern = glob2rx("*.tif"),
                                   full.names = TRUE))
  sub_stations <- crop(stations, dem)
  gI <- extract(dem, sub_stations, method = "bilinear")
  
  sub_stations <- as.data.frame(sub_stations)
  sub_stations$ELEV.M.COR <- gI
  
  # if original elevation == 0 and DEM indicates difference >= 15m, use DEM
  sub_stations[, 13] <- ifelse(sub_stations[, 9] == 0 &
                                 sub_stations[, 13] >= sub_stations[, 9] + 15,
                               sub_stations[, 9],
                               sub_stations[, 13])
  
  # if original elevation > 0 and DEM indicates difference <= / >= 50m, use DEM
  sub_stations[, 13] <- ifelse(sub_stations[, 9] > 0 &
                                 sub_stations[, 13] >= sub_stations[, 9] + 50 |
                                 sub_stations[, 13] >= sub_stations[, 9] - 50,
                               sub_stations[, 9],
                               sub_stations[, 13])
  
  # if DEM elevation == NA, use original elevation
  sub_stations[, 13] <- ifelse(is.na(sub_stations[, 13]) &
                                 !is.na(sub_stations[, 9]),
                               sub_stations[, 9], sub_stations[, 13])
  
  # if original elevation == NA, use DEM elevation
  sub_stations[, 13] <- ifelse(is.na(sub_stations[, 9]) & !is.na(sub_stations[, 13]),
                               sub_stations[, 9], sub_stations[, 13])
  
  cor_stations[[i]] <- sub_stations
  
}
```


# Combine list into one dataframe, check and write data to disk

```{r, echo=TRUE}

stations <- data.table::rbindlist(cor_stations)

summary(stations)

#devtools::use_data(stations, overwrite = TRUE)

```

# Figures and tables
<!-- insert graphs of uncorrected vs corrected elevation? Table of dropped stations? --->

# Notes

Users of these data should take into account the following (from the [NCDC website](http://www7.ncdc.noaa.gov/CDO/cdoselect.cmd?datasetabbv=GSOD&countryabbv=&georegionabbv=)): 

> "The following data and products may have conditions placed on their international commercial use. They can be used within the U.S. or for non-commercial international activities without restriction. The non-U.S. data cannot be redistributed for commercial purposes. Re-distribution of these data by others must provide this same notification." [WMO Resolution 40. NOAA Policy](http://www.wmo.int/pages/about/Resolution40.html)


# References

Hijmans, RJ, SJ Cameron, JL Parra, PG Jones, A Jarvis, 2005, Very High
Resolution Interpolated Climate Surfaces for Global Land Areas. *International
Journal of Climatology*. 25: 1965-1978.
[DOI:10.1002/joc.1276](http://dx.doi.org/10.1002/joc.1276)

Jarvis, A, HI Reuter, A Nelson, E Guevara, 2008, Hole-filled SRTM for the
globe Version 4, available from the CGIAR-CSI SRTM 90m Database
([http://srtm.csi.cgiar.org](http://srtm.csi.cgiar.org))

Jarvis, A, J Rubiano, A Nelson, A Farrow and M Mulligan, 2004, Practical use of
SRTM Data in the Tropics: Comparisons with Digital Elevation Models Generated
From Cartographic Data. Working Document no. 198. Cali, CO.
International Centre for Tropical Agriculture (CIAT): 32.

