---
title: "Fetch and Clean 'isd_history.csv' File"
author: "Adam H. Sparks"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r color, echo = FALSE, results='asis'}
# crayon needs to be explicitly activated in Rmd
options(crayon.enabled = TRUE)
# Hooks needs to be set to deal with outputs
# thanks to fansi logic
old_hooks <- fansi::set_knit_hooks(knitr::knit_hooks, 
                                   which = c("output", "message", "error"))
```

# Introduction

The "isd_history.csv" file details GSOD station metadata.
These data include the start and stop years used by {GSODR} to pre-check requests before querying the server for download and the country code used by {GSODR} when sub-setting for requests by country.
The following checks are performed on the raw data file before inclusion in {GSODR},

-   Check for valid lon and lat values;

    -   isd_history where latitude or longitude are `NA` or both 0 are removed leaving only properly georeferenced stations,

    -   isd_history where latitude is < -90˚ or > 90˚ are removed,

    -   isd_history where longitude is < -180˚ or > 180˚ are removed.

-   A new field, STNID, a concatenation of the USAF and WBAN fields, is added.

# Data Processing

## Set up workspace

```{r load_libs, echo=TRUE, message=FALSE, output=FALSE, warning=FALSE}
library("sessioninfo")
library("skimr")
library("countrycode")
library("data.table")
```

## Download and clean data

```{r download_NE_data, echo=TRUE, message=FALSE, warning=FALSE}
# download data
new_isd_history <- fread("https://www1.ncdc.noaa.gov/pub/data/noaa/isd-history.csv")
```

## Add/drop columns and save to disk

```{r clean_and_reformat, echo=TRUE, cache=FALSE}
# pad WBAN where necessary
new_isd_history[, WBAN := sprintf("%05d", WBAN)]

# add STNID column
new_isd_history[, STNID := paste(USAF, WBAN, sep = "-")]
setcolorder(new_isd_history, "STNID")
setnames(new_isd_history, "STATION NAME", "NAME")

# remove stations where LAT or LON is NA
new_isd_history <- na.omit(new_isd_history, cols = c("LAT", "LON"))

# remove extra columns
new_isd_history[, c("USAF", "WBAN", "ICAO") := NULL]
```

## Add country names based on FIPS

```{r merge-country, eval=TRUE, message=FALSE}
new_isd_history <-
  new_isd_history[setDT(countrycode::codelist), on = c("CTRY" = "fips")]

new_isd_history <- new_isd_history[, c(
  "STNID",
  "NAME",
  "LAT",
  "LON",
  "ELEV(M)",
  "CTRY",
  "STATE",
  "BEGIN",
  "END",
  "country.name.en",
  "iso2c",
  "iso3c"
)]

# clean data
new_isd_history[new_isd_history == -999] <- NA
new_isd_history[new_isd_history == -999.9] <- NA
new_isd_history <-
  new_isd_history[!is.na(new_isd_history$LAT) &
                    !is.na(new_isd_history$LON),]
new_isd_history <-
  new_isd_history[new_isd_history$LAT != 0 &
                    new_isd_history$LON != 0,]
new_isd_history <-
  new_isd_history[new_isd_history$LAT > -90 &
                    new_isd_history$LAT < 90,]
new_isd_history <-
  new_isd_history[new_isd_history$LON > -180 &
                    new_isd_history$LON < 180,]

# set colnames to upper case
names(new_isd_history) <- toupper(names(new_isd_history))
setnames(new_isd_history,
         old = "COUNTRY.NAME.EN",
         new = "COUNTRY_NAME")

# set country names to be upper case for easier internal verifications
new_isd_history[, COUNTRY_NAME := toupper(COUNTRY_NAME)]

# set key for joins when processing CSV files
setkeyv(new_isd_history, "STNID")
```

## Show changes from last release

```{r diff-codes}
# ensure we aren't using a locally installed dev version
install.packages("GSODR", repos = "https://cloud.r-project.org/")
load(system.file("extdata", "isd_history.rda", package = "GSODR"))

# select only the cols of interest
x <- names(isd_history)
new_isd_history <- new_isd_history[, ..x] 

(isd_diff <- diffobj::diffPrint(new_isd_history, isd_history))

rm(isd_history)

isd_history <- new_isd_history
```

## View and save the data

```{r view-and-save}
str(isd_history)

# write rda file to disk for use with GSODR package
save(isd_history,
     file = "../inst/extdata/isd_history.rda",
     compress = "bzip2")

save(isd_diff,
     file = "../inst/extdata/isd_diff.rda",
     compress = "bzip2")
```

# Notes

## NOAA policy

Users of these data should take into account the following (from the [NCEI website](https://www.ncei.noaa.gov/access/metadata/landing-page/bin/iso?id=gov.noaa.ncdc:C00516)):

> The following data and products may have conditions placed on their international commercial use. They can be used within the U.S. or for non-commercial international activities without restriction. The non-U.S. data cannot be redistributed for commercial purposes. Re-distribution of these data by others must provide this same notification. A log of IP addresses accessing these data and products will be maintained and may be made available to data providers.  
> For details, please consult: [WMO Resolution 40. NOAA Policy](https://community.wmo.int/resolution-40)

## R System Information

```{r system_information, echo=FALSE}
session_info()
```
