---
title: "Fetch, clean and correct altitude in GSOD isd_history.csv Data"
author: "Adam H. Sparks - Center for Crop Health, University of Southern Queensland"
date: "06-02-2016"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This script will fetch station data from the ftp server and clean up for
inclusion in package in /data/stations.rda for the GSODR package.

The following changes are made:

* Stations where latitude or longitude are NA or both 0 were removed

* Stations where latitude is < -90˚ or > 90˚ were removed

* Stations where longitude is < -180˚ or > 180˚ were removed

* A new field, STNID, a concatenation of the USAF and WBAN fields, was added

* Stations were checked against Natural Earth 1:10 ADM0 Cultural data, stations
not mapping in the isd-history reported country were dropped

* 90m hole-filled SRTM digital elevation (Jarvis *et al.* 2008) was used to
identify and correct/remove elevation errors in data for station locations
between -60˚ and 60˚ latitude. This applies to cases here where elevation was
missing in the reported values as well. In case the station reported an
elevation and the DEM does not, the station reported value is taken. For
stations beyond -60˚ and 60˚ latitude, the values are station reported values in
every instance for the 90m column.

# R Data Processing

## Set up workspace

```{r set up workspace, echo=TRUE, messages=FALSE}
dem_tiles <- list.files(path.expand("~/Data/CGIAR-CSI SRTM"), 
                        pattern = glob2rx("*.tif"), full.names = TRUE)
crs <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"
cor_stations <- list()
tf <- tempfile()
```

## Download from Natural Earth and NCDC

```{r download NE data, echo=TRUE}
# import Natural Earth cultural 1:10m data (last download 31/05/2016)
curl::curl_download("http://www.naturalearthdata.com/http//www.naturalearthdata.com/download/10m/cultural/ne_10m_admin_0_countries.zip",
                    destfile = tf)
NE <- unzip(tf)
NE <- raster::shapefile("./ne_10m_admin_0_countries.shp")
unlink(tf)

# download data
stations <- readr::read_csv(
  "ftp://ftp.ncdc.noaa.gov/pub/data/noaa/isd-history.csv",
  col_types = "ccccccddddd",
  col_names = c("USAF", "WBAN", "STN.NAME", "CTRY", "STATE", "CALL",
                "LAT", "LON", "ELEV.M", "BEGIN", "END"), skip = 1)

stations[stations == -999.9] <- NA
stations[stations == -999] <- NA

countries <- readr::read_table(
  "ftp://ftp.ncdc.noaa.gov/pub/data/noaa/country-list.txt")[-1, c(1, 3)]

# join with countrycode data (do we need this?)
countries <- dplyr::left_join(countries, countrycode::countrycode_data,
                              by = c(FIPS = "fips104"))
```

## Reformat and clean station data file from NCDC

```{r clean and reformat station data, echo=TRUE, cache=TRUE}
# clean data
stations <- stations[!is.na(stations$LAT) & !is.na(stations$LON), ]
stations <- stations[stations$LAT != 0 & stations$LON != 0, ]
stations <- stations[stations$LAT > -90 & stations$LAT < 90, ]
stations <- stations[stations$LON > -180 & stations$LON < 180, ]
stations$STNID <- paste(stations$USAF, stations$WBAN, sep = "-")

xy <- dplyr::left_join(stations, countries, by = c("CTRY" = "FIPS"))
```

## Check data for inconsistencies

GSOD data have some inconsistencies in them, some of this has been removed
above with filtering. Further filtering is used remove stations reporting
locations in countries that do not match the physical coordinates reported.
Using [Natural Earth Data 1:10 Cultural Data](http://www.naturalearthdata.com/downloads/10m-cultural-vectors/),
the stations reported countries are checked against the country in which the
coordinates map.

Also, reported elevation may differ from actual. Hijmans *et al.* (2005) created
their own digital elevation model using Jarvis *et al.* (2004) and [GTOPO30 data](https://lta.cr.usgs.gov/GTOPO30)
for areas where there was no SRTM data available (>60˚). Here only the
hole-filled SRTM data, V4 (Jarvis *et al.* 2008) was used for correction of
agroclimatology data (-60˚ to 60˚). Any incorrect station elevations beyond
these values were ignored in this data set. Stations with incorrect elevation
were identified using `raster::extract(x, y, buffer = 200, fun = mean)` so that
surrounding cells are also used to determine the elevation at that point, 
reducing the chances of over or underestimating in mountainous areas.
See Hijmans *et al.* (2005) for more detailed information on this methodology.

The hole-filled SRTM data is large enough that it won't all fit in-memory on
most desktop computers. Using tiles allows this process to run on a modest
machine with minimal effort but does take some time to loop through all of the
tiles.

Data can be downloaded from the [CGIAR-CSI's](http://csi.cgiar.org/WhtIsCGIAR_CSI.asp)
ftp server, [srtm.csi.cgiar.org](ftp://srtm.csi.cgiar.org),
using an FTP client to facilitate this next step.

```{r check station data for errors, echo=TRUE, cache=TRUE}
# quality check station locations for reported country and lat/lon position
# agreement

# create spatial object to check for location
xy <- as.data.frame(xy)
sp::coordinates(xy) <- ~ LON + LAT
sp::proj4string(xy) <- sp::CRS(crs)

# check for location in country
point_check <- sp::over(xy, NE)
point_check <- as.data.frame(point_check)
stn_discard <- point_check[point_check$FIPS %in% point_check$FIPS_10_ == FALSE, ]
nrow(stn_discard)
```
Zero observations (rows) in stations_discard, the data look good, no need to
remove any


```{r check elvation errors, echo=TRUE, cache=TRUE}
# create a spatial object for extracting elevation values using spatial points
stations <- as.data.frame(stations)
sp::coordinates(stations) <- ~ LON + LAT
sp::proj4string(stations) <- sp::CRS(crs)

for (i in dem_tiles) {
  
  # Load the DEM tile
  dem <- raster::raster(i)
  sub_stations <- raster::crop(stations, dem)
  
  # in some cases the DEM represents areas where there is no station
  # check for that here and if no stations, go on to next iteration
  if (is.null(sub_stations)) next
  
  # use a 200m buffer to extract elevation from the DEM
  SRTM_buffered <- raster::extract(dem, sub_stations, buffer = 200, fun = mean)
  
  # extract without using the buffer, strictly using 90m data
  SRTM_90m <- raster::extract(dem, sub_stations)
  
  # convert spatial object back to normal data frame and add new fields
  sub_stations <- as.data.frame(sub_stations)
  sub_stations$ELEV.M.SRTM.90m.BUFFER <- SRTM_buffered
  sub_stations$ELEV.M.SRTM.90m.NO_BUFFER <- SRTM_90m
  
  cor_stations[[i]] <- sub_stations
  rm(sub_stations)
}

stations <- as.data.frame(data.table::rbindlist(cor_stations))

# some stations occur in areas where DEM has no data
# use original station elevation in these cells
stations[, 13] <- ifelse(is.na(stations[, 13]), stations[, 9], stations[, 13])
stations[, 14] <- ifelse(is.na(stations[, 14]), stations[, 9], stations[, 14])

summary(stations)
```

# Figures

```{r ggplot, echo=FALSE, message=FALSE}
library(ggplot2)
library(ggalt)
```

```{r SRTM 90m vs Reported Elevation, echo=TRUE, fig.cap="GSOD Reported Elevation versus CGIAR-CSI SRTM Elevation", warning=FALSE}
ggplot(data = stations, aes(x = ELEV.M, y = ELEV.M.SRTM.90m.NO_BUFFER)) +
  geom_point(alpha = 0.4, size = 0.5)

```

```{r Buffered SRTM 90m vs Reported Elevation, echo=TRUE, fig.cap="GSOD Reported Elevation versus CGIAR-CSI SRTM Buffered Elevation", warning=FALSE}
ggplot(data = stations, aes(x = ELEV.M, y = ELEV.M.SRTM.90m.BUFFER)) +
  geom_point(alpha = 0.4, size = 0.5)
```

```{r SRTM 90m vs Buffered SRTM 90m, echo=TRUE, fig.cap="CGIAR-CSI SRTM Buffered Elevation versus CGIAR-CSI SRTM 90m"}
stations <- dplyr::mutate(stations, DIFFERENCE =
                            ELEV.M.SRTM.90m.NO_BUFFER - ELEV.M.SRTM.90m.BUFFER)
stations$DIFFERENCE[stations$DIFFERENCE >= -10 & stations$DIFFERENCE <= 10] <- NA

ggplot(data = stations, aes(x = LON, y = LAT)) +
  geom_point(aes(alpha = (DIFFERENCE)), size = 0.08) +
  labs(alpha = "NB - B") +
  coord_proj()
```

Assuming that a 10m difference in elevation between the buffered and unbuffered is
not important. Setting any values with a difference of >= -10 and <=10 to `NA`;
the results are still not geographically clustered in any one location or
feature.

```{r Hist difference SRTM 90m vs Buffered SRTM 90m, echo=TRUE, fig.cap="Histogram plot of CGIAR-CSI SRTM Buffered Elevation versus CGIAR-CSI SRTM 90m", warning=FALSE, error=FALSE, message=FALSE}
ggplot(data = stations) +
  geom_histogram(aes(ELEV.M.SRTM.90m.NO_BUFFER - ELEV.M.SRTM.90m.BUFFER))
```

The differences between the buffered and unbuffered elevation checks are minor
and appear to be normally distributed while also not showing any discernable
geographic pattern. However, The buffered elevation data are higher than the
unbuffered data. To help avoid within cell and between cell variation the
buffered values are the values that are included in the final data for
distribution with the GSODR package following the approach of Hijmans *et al.*
(2005). The new field is simply called ELEV.M.SRTM.90m in the stations.rda file.

```{r cleanup and .rda file creation, message=FALSE, echo=FALSE, warning=FALSE, error=FALSE}

# drop the 90m only station field
stations[, 14:15] <- NULL

# round SRTM.90m.Buffer field to whole number in cases where station reported
# data was used
stations[, 13] <- round(stations[, 13], 0)

names(stations)[13] <- "ELEV.M.SRTM.90m"

# write rda file to disk
devtools::use_data(stations, overwrite = TRUE, compress = "bzip2")

# clean up Natural Earth data files before we leave
file.remove(list.files(pattern = glob2rx("ne_10m_admin_0_countries*")))

```

# Notes

## NOAA Policy

Users of these data should take into account the following (from the [NCDC website](http://www7.ncdc.noaa.gov/CDO/cdoselect.cmd?datasetabbv=GSOD&countryabbv=&georegionabbv=)): 

> "The following data and products may have conditions placed on their international commercial use. They can be used within the U.S. or for non-commercial international activities without restriction. The non-U.S. data cannot be redistributed for commercial purposes. Re-distribution of these data by others must provide this same notification." [WMO Resolution 40. NOAA Policy](http://www.wmo.int/pages/about/Resolution40.html)

## R System Information

```{r system information, echo=FALSE}
sessionInfo()
```

# References

Hijmans, RJ, SJ Cameron, JL Parra, PG Jones, A Jarvis, 2005, Very High
Resolution Interpolated Climate Surfaces for Global Land Areas. *International
Journal of Climatology*. 25: 1965-1978.
[DOI:10.1002/joc.1276](http://dx.doi.org/10.1002/joc.1276)

Jarvis, A, HI Reuter, A Nelson, E Guevara, 2008, Hole-filled SRTM for the
globe Version 4, available from the CGIAR-CSI SRTM 90m Database
([http://srtm.csi.cgiar.org](http://srtm.csi.cgiar.org))

Jarvis, A, J Rubiano, A Nelson, A Farrow and M Mulligan, 2004, Practical use of
SRTM Data in the Tropics: Comparisons with Digital Elevation Models Generated
From Cartographic Data. Working Document no. 198. Cali, CO.
International Centre for Tropical Agriculture (CIAT): 32. [URL](http://srtm.csi.cgiar.org/PDF/Jarvis4.pdf)

