---
title: "Fetch GSOD Country List and Merge with ISO Country Codes"
author: "Adam H. Sparks"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Introduction

This script will fetch the country list provided by the NCDC for the GSOD
stations from the ftp server and merge it with ISO codes from the
[`countrycode`](https://cran.r-project.org/package=countrycode)
package for inclusion in the GSODR package in /data/country-list.rda. These
codes are used when a user selects a single country for a data query.

# R Data Processing

Read "country-list.txt" file from NCDC FTP server and merge with `countrycode`
data.

```{r download_merge_inspect, echo=TRUE, messages=FALSE}
if (!require("countrycode"))
{
  install.packages("countrycode")
}

countries <- readr::read_table(
  "ftp://ftp.ncdc.noaa.gov/pub/data/noaa/country-list.txt")[-1, c(1, 3)]
names(countries)[2] <- "COUNTRY_NAME"

GSOD_country_list <- dplyr::left_join(countries, countrycode::countrycode_data,
                   by = c(FIPS = "fips104"))
GSOD_country_list <- data.table::setDT(GSOD_country_list)

print(GSOD_country_list)
```

There are unnecessary data in several columns. `GSODR` only requires FIPS, name,
and ISO codes to function.

```{r remove_uncessesary, echo=TRUE, messages=FALSE}

GSOD_country_list[, c(3, 4:8, 11:16) := NULL]

print(GSOD_country_list)
```

Write .rda file to disk.

```{r write, echo=TRUE, messages=FALSE}
devtools::use_data(GSOD_country_list, overwrite = TRUE, compress = "bzip2")
```
# Notes

## NOAA Policy

Users of these data should take into account the following (from the [NCDC website](http://www7.ncdc.noaa.gov/CDO/cdoselect.cmd?datasetabbv=GSOD&countryabbv=&georegionabbv=)): 

> "The following data and products may have conditions placed on their international commercial use. They can be used within the U.S. or for non-commercial international activities without restriction. The non-U.S. data cannot be redistributed for commercial purposes. Re-distribution of these data by others must provide this same notification." [WMO Resolution 40. NOAA Policy](http://www.wmo.int/pages/about/Resolution40.html)

## R System Information

```{r system information, echo=FALSE}
sessionInfo()
```
