---
title: "Fetch, Clean and Correct Altitude in GSOD 'isd_history.csv' Data"
author: "Adam H. Sparks"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The isd_history file details station metadata including the start and stop years used by GSODR to pre-check requests before querying the server for download and the country code used by GSODR when subsetting for requests by country. The following changes are made to the raw data file for inclusion in _GSODR_:

  * isd_history where latitude or longitude are `NA` or both 0 are removed

  * isd_history where latitude is < -90˚ or > 90˚ are removed

  * isd_history where longitude is < -180˚ or > 180˚ are removed

  * A new field, STNID, a concatenation of the USAF and WBAN fields, is added

# Data Processing

## Set up workspace

```{r set_up_workspace, echo=TRUE, message=FALSE, output=FALSE, warning=FALSE}
if (!require("sessioninfo")) {
  install.packages("sessioninfo", repos = "https://cran.rstudio.com/")
}

if (!require("skimr")) {
  install.packages("skimr", repos = "https://cran.rstudio.com/")
}

if (!require("countrycode"))
{
  install.packages("countrycode",
                   repos = c(CRAN = "https://cran.rstudio.com"))
}

if (!require("data.table")) {
  install.packages("data.table", repos = "https://cran.rstudio.com/")
}
```

## Download and clean data

```{r download_NE_data, echo=TRUE, message=FALSE, warning=FALSE}
# download data
isd_history <- fread("https://www1.ncdc.noaa.gov/pub/data/noaa/isd-history.csv")

# clean data
isd_history[isd_history == -999] <- NA
isd_history[isd_history == -999.9] <- NA
isd_history <- isd_history[!is.na(isd_history$LAT) & !is.na(isd_history$LON), ]
isd_history <- isd_history[isd_history$LAT != 0 & isd_history$LON != 0, ]
isd_history <- isd_history[isd_history$LAT > -90 & isd_history$LAT < 90, ]
isd_history <- isd_history[isd_history$LON > -180 & isd_history$LON < 180, ]
```

## Add/drop columns and save to disk

```{r clean_and_reformat, echo=TRUE, cache=FALSE}
# add STNID column
isd_history[, STNID := paste(USAF, WBAN, sep = "-")]
setcolorder(isd_history, "STNID")
setnames(isd_history, "STATION NAME", "NAME")
setkey(isd_history, "STNID")

# drop stations not in GSOD data
isd_history[, STNID_len := nchar(STNID)]
isd_history <- subset(isd_history, STNID_len == 12)

# remove extra columns
isd_history[, c("USAF", "WBAN", "ICAO", "ELEV(M)", "STNID_len") := NULL]
```

## Add country names based on FIPS

```{r merge-country, eval=TRUE, message=FALSE}
isd_history <-
  isd_history[countrycode::codelist, on = c("CTRY" = "fips")]

isd_history <- isd_history[, c(
  "STNID",
  "NAME",
  "LAT",
  "LON",
  "CTRY",
  "STATE",
  "BEGIN",
  "END",
  "country.name.en",
  "iso2c",
  "iso3c"
)]

# set colnames to upper case
names(isd_history) <- toupper(names(isd_history))
setnames(
  isd_history,
  old = c("CTRY", "COUNTRY.NAME.EN"),
  new = c("FIPS", "COUNTRY_NAME")
)

# set country names to be upper case for easier internal verifications
isd_history[, COUNTRY_NAME := toupper(COUNTRY_NAME)]
```

## View and save the data

```{r view-and-save}
isd_history

# write rda file to disk for use with GSODR package
save(isd_history,
     file = "../inst/extdata/isd_history.rda",
     compress = "bzip2",
     version = 2)
```

# Notes

## NOAA Policy

Users of these data should take into account the following (from the
[NCEI website](http://www7.ncdc.noaa.gov/CDO/cdoselect.cmd?datasetabbv=GSOD&countryabbv=&georegionabbv=)): 

> "The following data and products may have conditions placed on their
international commercial use. They can be used within the U.S. or for
non-commercial international activities without restriction. The non-U.S. data
cannot be redistributed for commercial purposes. Re-distribution of these data
by others must provide this same notification."
[WMO Resolution 40. NOAA Policy](http://www.wmo.int/pages/about/Resolution40.html)

## R System Information

```{r system_information, echo=FALSE}
session_info()
```
