<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{GSODR vignette}
%\VignetteEncoding{UTF-8}
-->

# Introduction

The GSODR package provides the ability to interact with GSOD data using
spatial methods. The `get_GSOD` function allows for the data to be saved as
a GeoPackage file which can be read by most GIS software packages or in R using
R's GIS capabilities with contributed packages as well.

Following is an example of how you might download and save GSOD annual data for
a given country, Philippines in this example, and convert it into a KML file for
viewing in GoogleEarth. The second portion uses the same GeoPackage file to
import the data back into R and combine the GSOD data with the included CHELSA
data and plot the station temperatures for daily GSOD, average monthly GSOD and
CHELSA temperatures (1979-2013).

## Example - Download and plot data for a single country
Download data for Philippines for year 2010 and generate a spatial, year summary
file, PHL-2010.gpkg, in the user's home directory.

```{r example_1, eval=TRUE, message=FALSE, results='hide'}
library(GSODR)
get_GSOD(years = 2010, country = "Philippines", dsn = "~/",
         filename = "PHL-2010", GPKG = TRUE, max_missing = 5)
```

```{r example_1.1, eval=TRUE, echo=TRUE}
library(rgdal)
library(spacetime)
library(plotKML)

layers <- ogrListLayers(dsn = path.expand("~/PHL-2010.gpkg"))
pnts <- readOGR(dsn = path.expand("~/PHL-2010.gpkg"), layers[1])

# Plot results in Google Earth as a spacetime object:
pnts$DATE = as.Date(paste(pnts$YEAR, pnts$MONTH, pnts$DAY, sep = "-"))
row.names(pnts) <- paste("point", 1:nrow(pnts), sep = "")

tmp_ST <- STIDF(sp = as(pnts, "SpatialPoints"),
                time = pnts$DATE - 0.5,
                data = pnts@data[, c("TEMP", "STNID")],
                endTime = pnts$DATE + 0.5)

shape = "http://maps.google.com/mapfiles/kml/pal2/icon18.png"

kml(tmp_ST, dtime = 24 * 3600, colour = TEMP, shape = shape, labels = TEMP,
    file.name = "Temperatures_PHL_2010-2010.kml", folder.name = "TEMP")

system("zip -m Temperatures_PHL_2010-2010.kmz Temperatures_PHL_2010-2010.kml")
```

Compare the GSOD weather data from the Philippines with climatic data provided
by the GSODR package in the `GSOD_clim` data set.

```{r example_1.2, eval=TRUE, fig.cap="Comparison of GSOD daily values and average monthly values with CHELSA climate monthly values", fig.width=8}
library(dplyr)
library(ggplot2)
library(reshape2)

data(GSOD_clim)
cnames <- paste0("CHELSA_temp_", 1:12, "_1979-2013")
clim_temp <- GSOD_clim[GSOD_clim$STNID %in% pnts$STNID,
                       paste(c("STNID", cnames))]
clim_temp_df <- data.frame(STNID = rep(clim_temp$STNID, 12),
                           MONTHC = as.vector(sapply(1:12, rep,
                                                    times = nrow(clim_temp))), 
                           CHELSA_TEMP = as.vector(unlist(clim_temp[, cnames])))

pnts$MONTHC <- as.numeric(paste(pnts$MONTH))
temp <- left_join(pnts@data, clim_temp_df, by = c("STNID", "MONTHC"))

temp <- temp %>% 
  group_by(MONTH) %>% 
  mutate(AVG_DAILY_TEMP = round(mean(TEMP), 1))

df_melt <- na.omit(melt(temp[, c("STNID", "DATE", "CHELSA_TEMP", "TEMP", "AVG_DAILY_TEMP")],
                        id = c("DATE", "STNID")))

ggplot(df_melt, aes(x = DATE, y = value)) +
  geom_point(aes(color = variable), alpha = 0.2) +
  scale_x_date(date_labels = "%b") +
  ylab("Temperature (C)") +
  xlab("Month") +
  labs(colour = "") +
  scale_color_brewer(palette = "Dark2") +
  facet_wrap( ~ STNID)

```

# Notes

## Elevation Values

90 metre (90m) hole-filled SRTM digital elevation (Jarvis *et al.* 2008) was
used to identify and correct/remove elevation errors in data for station
locations between -60˚ and 60˚ latitude. This applies to cases here
where elevation was missing in the reported values as well. In case the
station reported an elevation and the DEM does not, the station reported
is taken. For stations beyond -60˚ and 60˚ latitude, the values are
station reported values in every instance. See
<https://github.com/adamhsparks/GSODR/blob/devel/data-raw/fetch_isd-history.md>
for more detail on the correction methods.

## WMO Resolution 40. NOAA Policy

*Users of these data should take into account the following (from the
[NCDC website](http://www7.ncdc.noaa.gov/CDO/cdoselect.cmd?datasetabbv=GSOD&countryabbv=&georegionabbv=)):*

> "The following data and products may have conditions placed on their 
international commercial use. They can be used within the U.S. or for
non-commercial international activities without restriction. The
non-U.S. data cannot be redistributed for commercial purposes.
Re-distribution of these data by others must provide this same
notification." [WMO Resolution 40. NOAA
Policy](http://www.wmo.int/pages/about/Resolution40.html)

# References
Stachelek, J. 2016. Using the Geopackage Format with R. 
URL: https://jsta.github.io/2016/07/14/geopackage-r.html
